Efficient Calculation of Interaction Features on Sparse Matrices

### 2015-08-07 paper/spares_interaction.tex
fixed indentation in algorithm
single sided
.
clarifying what interaction features are

\documentclass[twoside,11pt]{article}

@@

Introduction
Interaction features are a way of capturing correlations between features in a machine 
learning setting. They consist of products of combinations of features. This work 
describes a method for efficiently calculating second degree interaction features on a 
sparse matrix, and could be generalized to higher orders.

Consider the following matrix:

\[ \left( \begin{array}{cccc}
3 & 0 & 0 & 3 \\
0 & 4 & 2 & 0 \\
1 & 2 & 0 & 3 \end{array} \right)\]

If each row is an instance vector, then the interaction feature matrix is

\[ \left( \begin{array}{cccccc}
0 & 0 & 9 & 0 & 0 & 0 \\
0 & 0 & 0 & 8 & 0 & 0 \\
2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]

Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
since we generate products of all combinations of 2 features in the original matrix. 
Which column corresponds with which product pair is not important so long as it’s 
consistent. In this example, each column in the interaction matrix corresponded to the 
following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.

Notice that the interaction matrix contains many zero entries. This is of course because 
the original matrix contained zero entries, so the interaction features, which are 
products of pairs of features, contain many zeros. This means that the only products that 
need to actually be calculated or those for which both features in the combination are 
nonzero. If the original matrix is sparse and represented in a sparse matrix format 
(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
list via the following method.

\section{Approach}
Let the list of nonzero columns for a given row be $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to do this efficiently, a consistent mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from the space (a, b) where a,b are in $1,2,..., D$ onto the space $1,2,..., \frac{D^2-D}{2}$. This 
is isomorphic to mapping the coordinates of the upper triangle of a matrix onto a flat 
list. The following is a proof by construction for such a mapping.

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
\li $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
\li $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
\li \For $\id{row} in A$
\li     $N_{zc} \gets nonzero columns of row$
\li     \For each combination $a, b$ of elements of $N_{zc}$
\li         $k \gets \func{map}(b, a)$
\li         $i \gets index of \id{row}$
\li         $B[i, k] \gets \id{row}[a] * \id{row}[b]$
        \End
    \End
\end{codebox}

@@

\begin{align*}
N \binom{d D}{2} &= \frac{N (Dd)!}{2!(Dd-2)!}\\
    \\
    &= \frac{(D^2d^2-Dd)}{2}
\end{align*}

>>>>

\documentclass[11pt]{article}

@@

Introduction
Interaction features are a way of capturing correlations between features in a machine 
learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
$\{x_a \cdot x_b : a, b \in \{0,1,..., D-1\} \land a < b\}$, 
so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
approach to calculating these features is to simply iterate through the combinations of the column indices.
For a sparse vector, many of the resulting interaction features would be zero, and could therefore be ignored.
This work describes a method to efficiently calculate second degree interaction features for a sparse matrix 
that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.

%Consider the following matrix:

%\[ \left( \begin{array}{cccc}
%3 & 0 & 0 & 3 \\
%0 & 4 & 2 & 0 \\
%1 & 2 & 0 & 3 \end{array} \right)\]

%If each row is an instance vector, then the interaction feature matrix is

%\[ \left( \begin{array}{cccccc}
%0 & 0 & 9 & 0 & 0 & 0 \\
%0 & 0 & 0 & 8 & 0 & 0 \\
%2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]

%Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
%since we generate products of all combinations of 2 features in the original matrix. 
%Which column corresponds with which product pair is not important so long as it’s 
%consistent. In this example, each column in the interaction matrix corresponded to the 
%following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.

%Notice that the interaction matrix contains many zero entries. This is of course because 
%the original matrix contained zero entries, so the interaction features, which are 
%products of pairs of features, contain many zeros. This means that the only products that 
%need to actually be calculated or those for which both features in the combination are 
%nonzero. If the original matrix is sparse and represented in a sparse matrix format 
%(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
%and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
%list via the following method.

\section{Approach}
Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from the space (a, b) where a,b are in $1,2,..., D$ onto the space $1,2,..., \frac{D^2-D}{2}$. This 
is isomorphic to mapping the coordinates of the upper triangle of a matrix onto a flat 
list. The following is a proof by construction for such a mapping.

@@

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
    \zi $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
    \zi \For $\id{row}$ in $A$ \Do
    \zi     $N_{zc} \gets$ nonzero columns of $row$
    \zi     \For each combination $a, b$ of elements of $N_{zc}$ \Do
    \zi         $k \gets \func{map}(b, a)$
    \zi         $i \gets$ index of $\id{row}$
    \zi         $B[i, k] \gets \id{row}[a] \cdot \id{row}[b]$
            \End
       	\End
\end{codebox}

@@

\begin{align*}
N \binom{d D}{2} &= N \frac{(Dd)!}{2!(Dd-2)!}\\
    \\
    &= N \frac{(D^2d^2-Dd)}{2}
\end{align*}
